---
title: "PostHoc_weatherAnalysis"
format: pdf
editor: visual
---

## Post-hoc weather analysis: selecting weather variables

The complexity of our integrated model does not allow including an unlimited number of weather variables, and many weather variables are correlated to each other. Therefore, we have to select which one out of a candidate list will explain most of the variation. This document describes the process we followed to select the most relevant weather variable to include in the bigger integrated model.

NOTE: For the gyrfalcon, we tested several weather variables, but realised that there was not sufficient variation in the random effects. We still described the process in this document to show which variables we intended to test, but none them show correlations, which is entirely explained by the lack of variation. We therefore decided not to interpret these results in this part of the analysis.

To start, we load the necessary packages, relevant R-functions, and we set area names and years. We also load the model output of the complete model, without any weather covariates.

```{r}
#| label: Initial setup
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
library(tidyverse)
library(sf)
library(terra)
library(parallel)
library(coda)
library(brms)

## Set localities/areas and time period of interest
areas <- c("Hardangervidda", 
           "Dovrefjell", 
           "Børgefjell")
minYear <- 1991
maxYear <- 2020

## Source all functions in "R_weather" folder
sourceDir <- function(path, trace = TRUE, ...) {
  for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
    if(trace) cat(nm,":")
    source(file.path(path, nm), ...)
    if(trace) cat("\n")
  }
}
sourceDir('R_weather')

# Read in IDSM output
IDSM.out <- readRDS("/cloud/project/rypeIDSM_dHN_gyrData_23-10_mediumrun_fullloop_onlyOcc_RE.rds")
```

### Gyrfalcons

#### Extract and process weather data

For this analysis we need weather data, so in the code chunk below, we load those and process them so we can use this weather dataframe to test correlations later on. The wrangleData_GyrWeather function loads snowdepth, temperature and precipitation data from every day during the study period, and converts this to specific averages relevant to our study. The weather is extracted based on the centerpoints of gyrfalcon territories, and averaged per area. Raw weather data is extracted from SeNorge. The included averages are:

-   chicktemp: mean temperature during the nestling phase (between 15th of May (135) to 1st of July (182))

-   apriltemp: mean temperature during the early breeding phase (egg laying, April)

-   febtemp: mean temperature during territory establishment (February)

-   febsnow: mean snow depth during territory establishment (February)

-   SD20: mean snow depth during chick development (20th of May)

-   precip: sum of precipitation during the nestling period (between 15th of May (135) to 1st of July (182))

-   longrain: number of periods with 5 consecutive rainy days during the nestling period

```{r}
#| label: Load and process gyrfalcon weather data
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
## Read in and process weather data
d_weather <- wrangleData_GyrWeather(areas = areas,
                                    minYear = minYear, 
                                    maxYear = maxYear,
                                    byArea = FALSE)

# Define year range 
years <- seq_along(minYear:maxYear)

d_weather_long <- list()

# Loop through each element in the weather list
for (i in seq_along(d_weather)) {
  metric_name <- names(d_weather)[i]
  mat <- d_weather[[i]]$data
  
  # Convert to data frame
  df <- as.data.frame(mat)
  colnames(df) <- as.character(years)
  
  # Assign area index only if multiple rows (i.e., by_area = TRUE)
  if (nrow(df) > 1) {
    df$area <- seq_len(nrow(df))
  }
  
  # Reshape to long format
  df_long <- df %>%
    tidyr::pivot_longer(
      cols = if ("area" %in% colnames(df)) colnames(df)[colnames(df) != "area"] else everything(),
      names_to = "year",
      values_to = "value"
    ) %>%
    mutate(year = as.integer(year)) %>%
    rename(!!metric_name := value)
  
  # Store in list
  d_weather_long[[metric_name]] <- df_long
}

```

#### Extract and process random effects posteriors

Then, we extract posterior summaries (posterior mean and SD) for the random effects included in the models for gyrfalcon brood initiation and productivity.

```{r}
#| label: Extract and process gyr posterior summaries
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
## Extract posterior mean values and SD's for posthoc testing

# Occupancy/brood initiation rates
epsOcc_summary <- extractParamSummary(modelOutput = IDSM.out,
                                      parameterName = "epsT.Occ",
                                      byArea = FALSE)

epsOcc_summary <- epsOcc_summary %>%
  mutate(meanOcc = mean,
         sdOcc = sd) %>%
  select(-c(mean, sd))

# Productivity
epsProd_summary <- extractParamSummary(modelOutput = IDSM.out,
                                       parameterName = "epsT.Prod",
                                       byArea = FALSE)

epsProd_summary <- epsProd_summary %>%
  mutate(meanProd = mean,
         sdProd = sd) %>%
  select(-c(mean, sd))
```

Then we prep both datasets to be combined, and we set up a correlation matrix to identify weather variables that cannot be included in the same linear model.

```{r}
#| label: Data setup and correlation matrix gyrfalcon
#| echo: true
## Setup data for modelling

# Edit weather function to return means per year, not per year per area. 

residual_summary <- epsOcc_summary %>%
  left_join(epsProd_summary)

model_data <- residual_summary

for (df_name in names(d_weather_long)) {
  model_data <- model_data %>%
    left_join(d_weather_long[[df_name]], by = c("year"))
}


## Make correlation matrix

cor_matrix <- cor(model_data[, c("chicktemp", "apriltemp", "febtemp", "febsnow", "SD20", "precip", "longrain")], use = "complete.obs")

# Plot
library(ggcorrplot)
ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower",
           lab = TRUE, lab_size = 3, method = "square",
           colors = c("blue", "white", "red"))


```

The correlation matrix shows that some variables are strongly correlated. We will not include any of the correlated variables that show a correlation coefficient higher than 0.4 in the same model.

#### Modelling: *brms* linear models

Now we can run a series of linear models and plot their output to compare effect sizes and directions of correlations with brood initiation rates and productivity. For this, we used the *brms* package, and we used the default modelling settings.

**Model 1: occupancy/brood initiation rates and weather conditions in February**

```{r}
#| label: Occupancy model
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
# Occupancy
O1 <- brm(
  meanOcc | se(sdOcc) ~ febsnow + febtemp,
  data = model_data,
  family = gaussian()
)
```

```{r}
plot(O1)

rm(O1)
```

**Model 2: productivity and temperature + precipitation during the nestling period**

```{r}
#| label: Productivity model 1
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
P1 <- brm(
  meanProd | se(sdProd) ~ chicktemp + precip,
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: plot P1
plot(P1)
rm(P1)
```

**Model 3: productivity and temperature + the amount of long-rain events during the nestling period**

```{r}
#| label: Productivity model 2
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
P2 <- brm(
  meanProd | se(sdProd) ~ chicktemp + longrain,
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: plot P2
plot(P2)
rm(P2)
```

**Model 4: productivity and april temperature + precipitation during the nestling phase**

```{r}
#| label: Productivity model 3
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
P3 <- brm(
  meanProd | se(sdProd) ~ apriltemp + precip,
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: plot P3
plot(P3)
rm(P3)
```

As mentioned above, none of these variables show significant correlations with the posterior means of the random effects of gyrfalcon brood initiation rates or productivity. This analysis should be repeated with higher quality data sets to confidently draw conclusions about correlations with weather variables.

### **Ptarmigan**

The approach for the analysis for ptarmigan will be very similar to that of the gyrfalcon, except here we selected different weather variables, and we use the posterior means from the ptarmigan vital rates models on recruitment and survival. These did show sufficient variation for such a post-hoc analysis.

#### Extract and process weather data

Again, we start with extracting the weather data from SeNorge, and processing those data into the desired format. This time, weather is extracted from points along all transect lines in the dataset, spaced 500 meters apart. The wrangleData_Temp function samples these points, and extracts and averages the values for each weather variable. Because these take some time to run, the ready-made datasets are also saved in the repository and can simply be imported here to save time. Each of the averages are then formatted into the desired long format needed for our weather analysis. The weather variables we extracted are:

-   TempPost: temperature after incubation, during the chick rearing phase between 24 June to 15 July.

-   TempPre: temperature during the (pre-)incubation period between 1 May and 2 June.

-   TempJan: mean temperature in January, as a proxy for the harshness of the winters.

-   SnowFree: the first consecutive 7-day period where snow depth is under 1 cm

-   SD20: mean snow depth on 20 May

-   SnowPeriod: total length of the period in spring/summer where the snow depth is below 1 cm.

```{r}
#| label: Load and process ptarmigan weather data
#| echo: false
#| results: "hide"
#| message: false
#| warning: false
# Read in and process weather data

d_temp <- readRDS("data/weather/temp.rds") # Temperature during chick rearing

# d_temppre <- wrangleData_Temp(minYear = minYear,
#                               maxYear = maxYear,
#                               areas = areas,
#                               startday = 121,
#                               endday = 153)

d_temppre <- readRDS("data/weather/temppre.rds") # Temperature before and during incubation

# d_snow <- wrangleData_Snow(minYear = minYear,
#                            maxYear = maxYear,
#                            areas = areas)

d_snow <- readRDS("data/weather/snow.rds") # First 7-day snow free period

# d_snowdepth <- wrangleData_SnowDepth(minYear = minYear,
#                                      maxYear = maxYear,
#                                      areas = areas)

d_snowdepth <- readRDS("data/weather/snowdepth.rds") # Snow depth on the 20th of May

# d_tempJan <- wrangleData_Temp(minYear = minYear,
#                                  maxYear = maxYear,
#                                  areas = areas,
#                                  startday = 1,
#                                  endday = 31)

d_tempJan <- readRDS("data/weather/tempJan.rds") # Temperature in January

# d_snowperiod <- wrangleData_SnowPeriod(minYear = minYear,
#                                        maxYear = maxYear,
#                                        areas = areas)

d_snowperiod <- readRDS("data/weather/snowperiod.rds") # Total length of snow-free period

weather <- list(d_temp, d_temppre, d_tempJan, d_snow, d_snowdepth, d_snowperiod)

# Define year range 
years <- seq_along(minYear:maxYear)

# Initialize an empty list to store reshaped data frames
reshaped_list <- list()

# Define names for each metric (adjust to match the list structure)
metric_names <- c("TempPost", "TempPre", "TempJan", "SnowFree", "SD20", "SnowPeriod")

# Loop through each element in the weather list
for (i in seq_along(weather)) {
  # Choose which matrix to use: 'data' or 'standardized'
  if (!is.null(weather[[i]]$standardized)) {
    mat <- weather[[i]]$standardized
  } else {
    mat <- weather[[i]]$data
  }
  
  # Convert to data frame
  df <- as.data.frame(mat)
  colnames(df) <- as.character(years)
  df$area <- rownames(mat) %||% seq_len(nrow(df))  # fallback if no rownames
  
  # Reshape to long format
  df_long <- df %>%
    pivot_longer(
      cols = -area,
      names_to = "year",
      values_to = metric_names[i]
    ) %>%
    mutate(year = as.integer(year))
  
  # Store in list
  reshaped_list[[metric_names[i]]] <- df_long
}

```

#### Extract and process random effects posteriors

Then, we extract posterior summaries (posterior mean and SD) again, for the random effects included in the models for ptarmigan survival and recruitment.

```{r}
#| label: Extract and process posterior summaries ptarmigan
#| echo: false
#| results: "hide"
#| message: false
#| warning: false
## Extract posterior mean values and SD's for posthoc testing

# Ptarmigan recruitment
epsR_summary <- extractParamSummary(modelOutput = IDSM.out,
                                      parameterName = "epsR.R",
                                    byArea = TRUE)

epsR_summary <- epsR_summary %>%
  mutate(meanR = mean,
         sdR = sd) %>%
  select(-c(mean, sd))

# Ptarmigan survival
epsS_summary <- extractParamSummary(modelOutput = IDSM.out,
                                    parameterName = "epsR.S",
                                    byArea = TRUE)

epsS_summary <- epsS_summary %>%
  mutate(meanS = mean,
         sdS = sd) %>%
  select(-c(mean, sd))
```

Then we prep both data sets to be combined, and we set up a correlation matrix to identify weather variables that cannot be included in the same linear model.

```{r}
#| label: Set up model data and correlation matrix ptarmigan
#| echo: false

residual_summary <- epsR_summary %>%
  left_join(epsS_summary)

model_data <- residual_summary

for (df_name in names(reshaped_list)) {
  model_data <- model_data %>%
    left_join(reshaped_list[[df_name]], by = c("area", "year"))
}

model_data$area <- as.factor(model_data$area)


## Make correlation matrix

cor_matrix <- cor(model_data[, c("TempPost", "TempPre", "TempJan", "SnowFree", "SD20", "SnowPeriod")], use = "complete.obs")

# Plot
ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower",
           lab = TRUE, lab_size = 3, method = "square",
           colors = c("blue", "white", "red"))

```

We see a high correlation between the three snow depth variables, SnowFree, SnowPeriod and SD20. These can therefore not be included in the same model. Other corralation coefficients are all below 0.3 suggesting no issues with being included in the same model.

#### Modelling: *brms* linear models

Now we can run a series of linear models again and plot their output to compare effect sizes and directions of correlations with brood initiation rates and productivity. For this, we used the *brms* package, and we used the default modelling settings.

##### Recruitment

We start with recruitment. We tested whether ptarmigan recruitment was related to temperature and spring timing. Many studies report positive relationships between ptarmigan recruitment and spring air temperatures, which is also reflected in earlier snowmelt (46, 69-72). Generally, in years with warm springs, nutritious and fresh vegetation is available to ptarmigan and their offspring earlier and in higher abundance (73). For temperature, we calculated the average during the (pre-)incubation period between the 1 May and 2 June, and the average during the chick rearing period from 24 June to 15 July. For spring timing, we quantified the snow conditions by calculating the average snow depth on 20 May and the first 7-day period where snow depth is below 1 cm.

**Model 1: Recruitment and early spring temperature + late spring temperature + spring snow depth**

```{r}
#| label: Model recruitment 1
#| echo: true
#| results: "hide"
#| message: false
#| warning: false

R1 <- brm(
  meanR | se(sdR) ~ TempPre + TempPost + SD20, 
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: Plot R1
plot(R1)
rm(R1)
```

**Model 2: Recruitment and early spring temperature + late spring temperature**

```{r}
#| label: Model recruitment 2
#| echo: true
#| results: "hide"
#| message: false
#| warning: false

R2 <- brm(
  meanR | se(sdR) ~ TempPre + TempPost, 
  data = model_data,
  family = gaussian()
)


```

```{r}
#| label: Plot R2
plot(R2)
rm(R2)
```

**Model 3: recruitment and early spring temperature + length of the snowfree period**

```{r}
#| label: Model recruitment 3
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
R3 <- brm(
  meanR | se(sdR) ~ TempPre + SnowFree, 
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: Plot R3
plot(R3)
rm(R3)
```

This shows that TempPre, TempPost and SnowFree all have positive effects on recruitment. Based on these figures, we can conclude that early spring temperature always shows a positive effect that is stronger in effect size compared to the other weather variables that are related to recruitment.

##### Survival

We tested whether ptarmigan survival was related to temperature and snow conditions throughout the year. Previous studies show that snow depth in the spring is negatively related to ptarmigan survival (7, 8). We used the same spring snow depth index as above, as the average snow depth on 20 May. We also calculated the average length of the snow-free period in the summer, as the number of consecutive days where snow depth is below 1 cm. We calculated the mean snow depth during the pre-breeding period, which we defined as the period between 1 May and 2 June, which includes ptarmigan laying and pre-laying days.

**Model 1: survival and early spring temperature + winter temperature + length of the snowfree period**

```{r}
#| label: Model survival 1
#| echo: true
#| results: "hide"
#| message: false
#| warning: false

S1 <- brm(
  meanS | se(sdS) ~ TempPre + TempJan + SnowPeriod, 
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: Plot S1
plot(S1)
rm(S1)
```

**Model 2: survival and early spring temperature + winter temperature + spring snow depth**

```{r}
#| label: Model survival 2
#| echo: true
#| results: "hide"
#| message: false
#| warning: false

S2 <- brm(
  meanS | se(sdS) ~ TempPre + TempJan + SD20, 
  data = model_data,
  family = gaussian()
)

```

```{r}
#| label: Plot S2
plot(S2)
rm(S2)
```

**Model 3: Survival and winter temperature + length of the snow free period**

```{r}
#| label: Model survival 3
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
S3 <- brm(
  meanS | se(sdS) ~ TempJan + SnowPeriod, 
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: Plot S3
plot(S3)
rm(S3)
```

**Model 4: survival and early spring temperature + length of the snow free period**

```{r}
#| label: Model survival 4
#| echo: true
#| results: "hide"
#| message: false
#| warning: false
S4 <- brm(
  meanS | se(sdS) ~ TempPre + SnowPeriod, 
  data = model_data,
  family = gaussian()
)
```

```{r}
#| label: Plot S4
plot(S4)
rm(S4)
```

The models based on ptarmigan survival show that ... Elaborate!

### Conclusions

Based on these analyses, we see that TempPre has the strongest and most consistent support from our post-hoc modelling analysis. Besides, it receives strong support from previous studies investigating the effect of weather on ptarmigan vital rates. Therefore, we decided to include temperature during the (pre-)incubation period as the main weather variable in our integrated model. Because it is strongly related to ptarmigan recruitment, it will be included in the model estimating recruitment.
