---
title: "PostHoc_weatherAnalysis"
format: html
editor: visual
---

## Post-hoc weather analysis: selecting weather variables

The complexity of our integrated model does not allow including an unlimited number of weather variables, and many weather variables are correlated to each other. Therefore, we have to select which one out of a candidate list will explain most of the variation. This document describes the process we followed to select the most relevant weather variable to include in the bigger integrated model.

NOTE: For the gyrfalcon, we tested several weather variables, but realised that there was not sufficient variation in the random effects. We still described the process in this document to show which variables we intended to test, but none them show correlations, which is entirely explained by the lack of variation. We therefore decided not to interpret these results in this part of the analysis.

To start, we load the necessary packages, relevant R-functions, and we set area names and years. We also load the model output of the complete model, without any weather covariates.

```{r}
#| label: Initial setup
#| echo: false
library(tidyverse)
library(sf)
library(terra)
library(parallel)
library(coda)
library(brms)

## Set localities/areas and time period of interest
areas <- c("Hardangervidda", 
           "Dovrefjell", 
           "Børgefjell")
minYear <- 1991
maxYear <- 2020

## Source all functions in "R_weather" folder
sourceDir <- function(path, trace = TRUE, ...) {
  for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
    if(trace) cat(nm,":")
    source(file.path(path, nm), ...)
    if(trace) cat("\n")
  }
}
sourceDir('R_weather')

# Read in IDSM output
IDSM.out <- readRDS("/cloud/project/rypeIDSM_dHN_gyrData_23-10_mediumrun_fullloop_onlyOcc_RE.rds")
```

### Gyrfalcons

#### Extract and process weather data

For this analysis we need weather data, so in the code chunk below, we load those and process them so we can use this weather dataframe to test correlations later on. The wrangleData_GyrWeather function loads snowdepth, temperature and precipitation data from every day during the study period, and converts this to specific averages relevant to our study. The weather is extracted based on the centerpoints of gyrfalcon territories, and averaged per area. Raw weather data is extracted from SeNorge. The included averages are:

-   chicktemp: mean temperature during the nestling phase (between 15th of May (135) to 1st of July (182))

-   apriltemp: mean temperature during the early breeding phase (egg laying, April)

-   febtemp: mean temperature during territory establishment (February)

-   febsnow: mean snow depth during territory establishment (February)

-   SD20: mean snow depth during chick development (20th of May)

-   precip: sum of precipitation during the nestling period (between 15th of May (135) to 1st of July (182))

-   longrain: number of periods with 5 consecutive rainy days during the nestling period

```{r}
#| label: Load and process gyrfalcon weather data
#| echo: false
## Read in and process weather data
d_weather <- wrangleData_GyrWeather(areas = areas,
                                    minYear = minYear, 
                                    maxYear = maxYear,
                                    byArea = FALSE)

# Define year range 
years <- seq_along(minYear:maxYear)

d_weather_long <- list()

# Loop through each element in the weather list
for (i in seq_along(d_weather)) {
  metric_name <- names(d_weather)[i]
  mat <- d_weather[[i]]$data
  
  # Convert to data frame
  df <- as.data.frame(mat)
  colnames(df) <- as.character(years)
  
  # Assign area index only if multiple rows (i.e., by_area = TRUE)
  if (nrow(df) > 1) {
    df$area <- seq_len(nrow(df))
  }
  
  # Reshape to long format
  df_long <- df %>%
    tidyr::pivot_longer(
      cols = if ("area" %in% colnames(df)) colnames(df)[colnames(df) != "area"] else everything(),
      names_to = "year",
      values_to = "value"
    ) %>%
    mutate(year = as.integer(year)) %>%
    rename(!!metric_name := value)
  
  # Store in list
  d_weather_long[[metric_name]] <- df_long
}

```

#### Extract and process random effects posteriors

Then, we extract posterior summaries (posterior mean and SD) for the random effects included in the models for gyrfalcon brood initiation and productivity.

```{r}
#| label: Extract and process gyr posterior summaries
#| echo: false
## Extract posterior mean values and SD's for posthoc testing

# Occupancy/brood initiation rates
epsOcc_summary <- extractParamSummary(modelOutput = IDSM.out,
                                      parameterName = "epsT.Occ",
                                      byArea = FALSE)

epsOcc_summary <- epsOcc_summary %>%
  mutate(meanOcc = mean,
         sdOcc = sd) %>%
  select(-c(mean, sd))

# Productivity
epsProd_summary <- extractParamSummary(modelOutput = IDSM.out,
                                       parameterName = "epsT.Prod",
                                       byArea = FALSE)

epsProd_summary <- epsProd_summary %>%
  mutate(meanProd = mean,
         sdProd = sd) %>%
  select(-c(mean, sd))
```

Then we prep both datasets to be combined, and we set up a correlation matrix to identify weather variables that cannot be included in the same linear model.

```{r}
#| label: Data setup and correlation matrix gyrfalcon
#| echo: false
## Setup data for modelling

# Edit weather function to return means per year, not per year per area. 

residual_summary <- epsOcc_summary %>%
  left_join(epsProd_summary)

model_data <- residual_summary

for (df_name in names(d_weather_long)) {
  model_data <- model_data %>%
    left_join(d_weather_long[[df_name]], by = c("year"))
}


## Make correlation matrix

cor_matrix <- cor(model_data[, c("chicktemp", "apriltemp", "febtemp", "febsnow", "SD20", "precip", "longrain")], use = "complete.obs")

# Plot
library(ggcorrplot)
ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower",
           lab = TRUE, lab_size = 3, method = "square",
           colors = c("blue", "white", "red"))


```

The correlation matrix shows that some variables are strongly correlated. We will not include both precipitation indices in the same model and we cannot put SD20 and febsnow in the same model. Other correlation coefficients are below 0.4 and should not pose too many problems during modelling.

#### Modelling: *brms* linear models

Now we can run a series of linear models and plot their output to compare effect sizes and directions of correlations with brood initiation rates and productivity. For this, we used the *brms* package, and we used the default modelling settings.

Model 1: occupancy/brood initiation rates and weather conditions in February

```{r}
#| label: Occupancy model
# Occupancy
O1 <- brm(
  meanOcc | se(sdOcc) ~ febsnow + febtemp,
  data = model_data,
  family = gaussian()
)

plot(O1)
summary(O1)
```

Model 2: productivity and temperature + precipitation during the nestling period

```{r}
#| label: Productivity model 1
P1 <- brm(
  meanProd | se(sdProd) ~ chicktemp + precip,
  data = model_data,
  family = gaussian()
)

plot(P1)
summary(P1)
```

Model 3: productivity and temperature + the amount of long-rain events during the nestling period

```{r}
#| label: Productivity model 2
P2 <- brm(
  meanProd | se(sdProd) ~ chicktemp + longrain,
  data = model_data,
  family = gaussian()
)

plot(P2)
summary(P2)
```

Model 4: productivity and april temperature + precipitation during the nestling phase

```{r}
#| label: Productivity model 3
P3 <- brm(
  meanProd | se(sdProd) ~ apriltemp + precip,
  data = model_data,
  family = gaussian()
)

plot(P3)
summary(P3)
```

As mentioned above, none of these variables show significant correlations with the posterior means of the random effects of gyrfalcon brood initiation rates or productivity. This analysis should be repeated with higher quality data sets to confidently draw conclusions about correlations with weather variables.

### Ptarmigan

The approach for the analysis for ptarmigan will be very similar to that of the gyrfalcon, except here we selected different weather variables, and we use the posterior means from the ptarmigan vital rates models on recruitment and survival. These did show sufficient variation for such a post-hoc analysis.

#### Extract and process weather data

Again, we start with extracting the weather data from SeNorge, and processing those data into the desired format. This time, weather is extracted from points along all transect lines in the dataset, spaced 500 meters apart. The wrangleData_Temp function samples these points, and extracts and averages the values for each weather variable. Because these take some time to run, the ready-made datasets are also saved in the repository and can simply be imported here to save time. Each of the averages are then formatted into the desired long format needed for our weather analysis. The weather variables we extracted are:

-   TempPost: temperature after incubation, during the chick rearing phase between 24 June to 15 July.

-   TempPre: temperature during the (pre-)incubation period between 1 May and 2 June.

-   TempJan: mean temperature in January, as a proxy for the harshness of the winters.

-   SnowFree: the first consecutive 7-day period where snow depth is under 1 cm

-   SD20: mean snow depth on 20 May

-   SnowPeriod: total length of the period in spring/summer where the snow depth is below 1 cm.

```{r}
#| label: Load and process ptarmigan weather data
# Read in and process weather data

d_temp <- readRDS("data/weather/temp.rds") # Temperature during chick rearing

# d_temppre <- wrangleData_Temp(minYear = minYear,
#                               maxYear = maxYear,
#                               areas = areas,
#                               startday = 121,
#                               endday = 153)

d_temppre <- readRDS("data/weather/temppre.rds") # Temperature before and during incubation

# d_snow <- wrangleData_Snow(minYear = minYear,
#                            maxYear = maxYear,
#                            areas = areas)

d_snow <- readRDS("data/weather/snow.rds") # First 7-day snow free period

# d_snowdepth <- wrangleData_SnowDepth(minYear = minYear,
#                                      maxYear = maxYear,
#                                      areas = areas)

d_snowdepth <- readRDS("data/weather/snowdepth.rds") # Snow depth on the 20th of May

# d_tempJan <- wrangleData_Temp(minYear = minYear,
#                                  maxYear = maxYear,
#                                  areas = areas,
#                                  startday = 1,
#                                  endday = 31)

d_tempJan <- readRDS("data/weather/tempJan.rds") # Temperature in January

# d_snowperiod <- wrangleData_SnowPeriod(minYear = minYear,
#                                        maxYear = maxYear,
#                                        areas = areas)

d_snowperiod <- readRDS("data/weather/snowperiod.rds") # Total length of snow-free period

weather <- list(d_temp, d_temppre, d_tempJan, d_snow, d_snowdepth, d_snowperiod)

# Define year range 
years <- seq_along(minYear:maxYear)

# Initialize an empty list to store reshaped data frames
reshaped_list <- list()

# Define names for each metric (adjust to match the list structure)
metric_names <- c("TempPost", "TempPre", "TempJan", "SnowFree", "SD20", "SnowPeriod")

# Loop through each element in the weather list
for (i in seq_along(weather)) {
  # Choose which matrix to use: 'data' or 'standardized'
  if (!is.null(weather[[i]]$standardized)) {
    mat <- weather[[i]]$standardized
  } else {
    mat <- weather[[i]]$data
  }
  
  # Convert to data frame
  df <- as.data.frame(mat)
  colnames(df) <- as.character(years)
  df$area <- rownames(mat) %||% seq_len(nrow(df))  # fallback if no rownames
  
  # Reshape to long format
  df_long <- df %>%
    pivot_longer(
      cols = -area,
      names_to = "year",
      values_to = metric_names[i]
    ) %>%
    mutate(year = as.integer(year))
  
  # Store in list
  reshaped_list[[metric_names[i]]] <- df_long
}

```

#### Extract and process random effects posteriors

Then, we extract posterior summaries (posterior mean and SD) again, for the random effects included in the models for ptarmigan survival and recruitment.

```{r}
#| label: Extract and process posterior summaries ptarmigan
## Extract posterior mean values and SD's for posthoc testing

# Ptarmigan recruitment
epsR_summary <- extractParamSummary(modelOutput = IDSM.out,
                                      parameterName = "epsR.R",
                                    byArea = TRUE)

epsR_summary <- epsR_summary %>%
  mutate(meanR = mean,
         sdR = sd) %>%
  select(-c(mean, sd))

# Ptarmigan survival
epsS_summary <- extractParamSummary(modelOutput = IDSM.out,
                                    parameterName = "epsR.S",
                                    byArea = TRUE)

epsS_summary <- epsS_summary %>%
  mutate(meanS = mean,
         sdS = sd) %>%
  select(-c(mean, sd))
```

Then we prep both data sets to be combined, and we set up a correlation matrix to identify weather variables that cannot be included in the same linear model.

```{r}
#| label: Set up model data and correlation matrix ptarmigan
residual_summary <- epsR_summary %>%
  left_join(epsS_summary)

model_data <- residual_summary

for (df_name in names(reshaped_list)) {
  model_data <- model_data %>%
    left_join(reshaped_list[[df_name]], by = c("area", "year"))
}

model_data$area <- as.factor(model_data$area)


## Make correlation matrix

cor_matrix <- cor(model_data[, c("TempPost", "TempPre", "TempJan", "SnowFree", "SD20", "SnowPeriod")], use = "complete.obs")

# Plot
ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower",
           lab = TRUE, lab_size = 3, method = "square",
           colors = c("blue", "white", "red"))

```

We see a high correlation between the three snow depth variables, SnowFree, SnowPeriod and SD20. These can therefore not be included in the same model. Other corralation coefficients are all below 0.3 suggesting no issues with being included in the same model.

#### Modelling: *brms* linear models

Now we can run a series of linear models again and plot their output to compare effect sizes and directions of correlations with brood initiation rates and productivity. For this, we used the *brms* package, and we used the default modelling settings. We included area as a fixed effect to account for repeated measurements from the same area.

##### Recruitment

We start with recruitment. We tested whether ptarmigan recruitment was related to temperature and spring timing. Many studies report positive relationships between ptarmigan recruitment and spring air temperatures, which is also reflected in earlier snowmelt (46, 69-72). Generally, in years with warm springs, nutritious and fresh vegetation is available to ptarmigan and their offspring earlier and in higher abundance (73). For temperature, we calculated the average during the (pre-)incubation period between the 1 May and 2 June, and the average during the chick rearing period from 24 June to 15 July. For spring timing, we quantified the snow conditions by calculating the average snow depth on 20 May and the first 7-day period where snow depth is below 1 cm.

```{r}
#| label: Modelling for ptarmigan recruitment
R1 <- brm(
  meanR | se(sdR) ~ TempPre + TempPost + SnowFree + (1 | area),
  data = model_data,
  family = gaussian()
)

R2 <- brm(
  meanR | se(sdR) ~ TempPre + TempPost + SD20 + (1 | area),
  data = model_data,
  family = gaussian()
)

R3 <- brm(
  meanR | se(sdR) ~ TempPre + TempPost + SnowFree + area,
  data = model_data,
  family = gaussian()
)

R4 <- brm(
  meanR | se(sdR) ~ TempPre + TempPost + area,
  data = model_data,
  family = gaussian()
)

plot(R4)

R5 <- brm(
  meanR | se(sdR) ~ TempPre + SnowFree + area,
  data = model_data,
  family = gaussian()
)

plot(R5)

R6 <- brm(
  meanR | se(sdR) ~ SnowFree + area,
  data = model_data,
  family = gaussian()
)

R7 <- brm(
  meanR | se(sdR) ~ TempPre + area,
  data = model_data,
  family = gaussian()
)
```

We tried several model setups and compared them using leave-one-out cross validation

```{r}
#| label: Model comparison ptarmigan recruitment
loo_R1 <- loo(R1)
loo_R2 <- loo(R2)
loo_R3 <- loo(R3)
loo_R4 <- loo(R4)
loo_R5 <- loo(R5)
loo_R6 <- loo(R6)
loo_R7 <- loo(R7)
loo_compare(loo_R1, loo_R2, loo_R3, loo_R4, loo_R6, loo_R5, loo_R7)
```

Based on these analyses, TempPre is always included and shows the strongest correlation with ptarmigan recruitment

##### Survival

We tested whether ptarmigan survival was related to temperature and snow conditions throughout the year. Previous studies show that snow depth in the spring is negatively related to ptarmigan survival (7, 8). We used the same spring snow depth index as above, as the average snow depth on 20 May. We also calculated the average length of the snow-free period in the summer, as the number of consecutive days where snow depth is below 1 cm. We calculated the mean snow depth during the pre-breeding period, which we defined as the period between 1 May and 2 June, which includes ptarmigan laying and pre-laying days.

```{r}
#| label: Modelling for ptarmigan survival
# Survival

S1 <- brm(
  meanS | se(sdS) ~ TempPre + TempJan + SnowPeriod + area,
  data = model_data,
  family = gaussian()
)

S2 <- brm(
  meanS | se(sdS) ~ TempPre + TempJan + SD20 + area,
  data = model_data,
  family = gaussian()
)

S3 <- brm(
  meanS | se(sdS) ~ TempJan + SnowPeriod + area,
  data = model_data,
  family = gaussian()
)

plot(S3)

S4 <- brm(
  meanS | se(sdS) ~ TempPre + SnowPeriod + area,
  data = model_data,
  family = gaussian()
)

plot(S4)

S5 <- brm(
  meanS | se(sdS) ~ TempPre + area,
  data = model_data,
  family = gaussian()
)

S6 <- brm(
  meanS | se(sdS) ~ TempJan + area,
  data = model_data,
  family = gaussian()
)

S7 <- brm(
  meanS | se(sdS) ~ SnowPeriod + area,
  data = model_data,
  family = gaussian()
)

```

Again, we set up several models and compared them using leave-one-out cross validation

```{r}
#| label: Model comparison ptarmigan survival
loo_S1 <- loo(S1)
loo_S2 <- loo(S2)
loo_S3 <- loo(S3)
loo_S4 <- loo(S4)
loo_S5 <- loo(S5)
loo_S6 <- loo(S6)
loo_S7 <- loo(S7)
loo_compare(loo_S1, loo_S2, loo_S3, loo_S4, loo_S5, loo_S6, loo_S7)
```

#### Plotting model results

After all the modelling, we want to visualise the results, to be able to select the most relevant weather variable.

```{r}
#| label: Plotting model results
## Plot effects

library(dplyr)
library(tidyr)
library(stringr)

# Example for one model
samples_R3 <- posterior_samples(R3, pars = "^b_")
samples_R3 <- samples_R3 %>%
  pivot_longer(cols = everything(), names_to = "param", values_to = "value") %>%
  mutate(model = "R3")

samples_S1 <- posterior_samples(S1, pars = "^b_")
samples_S1 <- samples_S1 %>%
  pivot_longer(cols = everything(), names_to = "param", values_to = "value") %>%
  mutate(model = "S1")

samples_S6 <- posterior_samples(S6, pars = "^b_")
samples_S6 <- samples_S6 %>%
  pivot_longer(cols = everything(), names_to = "param", values_to = "value") %>%
  mutate(model = "S6")

samples_S7 <- posterior_samples(S7, pars = "^b_")
samples_S7 <- samples_S7 %>%
  pivot_longer(cols = everything(), names_to = "param", values_to = "value") %>%
  mutate(model = "S7")

effects_df <- bind_rows(
  samples_R3 %>% mutate(model = "Recruitment 3"),
  samples_S1 %>% mutate(model = "Survival 1")
  # samples_S6 %>% mutate(model = "Survival 6"),
  # samples_S7 %>% mutate(model = "Survival 7")
)

effects_df <- effects_df %>%
  mutate(
    type = str_replace(param, "b_", "")
  )


summary_df <- effects_df %>%
  group_by(model, type) %>%
  summarise(
    median = median(value),
    lci = quantile(value, 0.025),
    uci = quantile(value, 0.975),
    .groups = "drop"
  ) %>%
  filter(type %in% c("SnowFree", "SnowPeriod", "TempJan", "TempPre", "TempPost"))



library(ggplot2)

ggplot(summary_df, aes(x = median, y = type)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lci, xmax = uci), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ model) +
  theme_minimal() +
  labs(x = "Posterior effect (median ± 95% CI)", y = "Covariate")
```

### Conclusions

Based on these analyses, we see that TempPre has the strongest and most consistent support from our post-hoc modelling analysis. Besides, it receives strong support from previous studies investigating the effect of weather on ptarmigan vital rates. Therefore, we decided to include temperature during the (pre-)incubation period as the main weather variable in our integrated model. Because it is strongly related to ptarmigan recruitment, it will be included in the model estimating recruitment.
